Resolvable Ambiguity

### 2019-07-10 paper/paper.tex
Pass over case studies

We have implemented a tool and a small DSL for syntax definition
that generates a language definition in the style of
Section~\ref{sec:parse-time-disambiguation}, which the tool then
uses to construct a parser and do a dynamic resolvability
analysis. The tool is implemented in Haskell and uses an
off-the-shelf
implementation\footnote{\url{http://hackage.haskell.org/package/Earley}}
of the Earley parsing algorithm
\cite{earleyEfficientContextfreeParsing1970}. The purpose of these
case studies is to showcase possible use cases of resolvable
ambiguity in the context of a pre-existing general purpose
programming language (OCaml, Section~\ref{sec:evaluation-ocaml}),
a pre-existing DSL (Orc, Section~\ref{sec:evaluation-orc}), and a
new DSL that is under development (Cyphym, Section~\ref{sec:evaluation-cyphym}).

The syntax definition DSL builds on the the ``syncons''
introduced by
\citet{palmkvistCreatingDomainSpecificLanguages2019}, wherein each
production is defined separately from each other (one
\syncon{syncon} each), and marks are primarily introduced through
precedence lists. For example, the running example used in
Section~\ref{sec:parse-time-disambiguation}
(Fig.~\ref{fig:running-example-definition} on
page~\pageref{fig:running-example-definition}) can be written as
follows:

\begin{tabular}{ll}
\begin{lstlisting}[language=syncon,boxpos=t]
type Exp
grouping "(" Exp ")"
precedence {
  mul; // higher in precedence list
  add; // means higher precedence
}
\end{lstlisting}
&
\begin{lstlisting}[language=syncon,boxpos=t]
token Integer = "[0-9]+"
syncon list: Exp =
  "[" (head:Exp (";" tail:Exp))? "]"
infix add: Exp = "+"
infix mul: Exp = "*"
syncon literal: Exp = n:Integer
\end{lstlisting}
\end{tabular}\smallskip

\noindent The design of the DSL diverges from the formalism in
Section~\ref{sec:parse-time-disambiguation} partially for
familiarity (precedence instead of marks), but also for
composability reasons: the DSL allows marks to be introduced
without altering the original definition of a syncon. If marks are
needed in non-precedence related situations, e.g., adding
sequential composition to the running example, they can be
introduced with ``forbids'':

\begin{lstlisting}[language=syncon]
infix seqComp: Exp = ";"
forbid list.head = seqComp
forbid list.tail = seqComp
\end{lstlisting}

\subsection{OCaml} \label{sec:evaluation-ocaml}

\subsection{Orc} \label{sec:evaluation-orc}
%TODO Title should be descriptive

In this section, we consider the use case of defining a language
and composing it with another previously defined language, showing
how we can deal with the resulting ambiguities that arise from the
composition.
%
The language being defined is a subset of
Orc~\cite{kitchinOrc2009}, a functional programming language which
includes a number of special-purpose combinators for coordinating
concurrent workflows.

@@

\subsection{Cyphym} \label{sec:evaluation-cyphym}

>>>>

We have implemented a tool and a small DSL for syntax definition
that generates a language definition in the style of
Section~\ref{sec:parse-time-disambiguation}, which the tool then
uses to construct a parser and do dynamic resolvability analysis.
The tool is implemented in Haskell and uses an off-the-shelf
implementation\footnote{\url{http://hackage.haskell.org/package/Earley}}
of the Earley parsing algorithm
\cite{earleyEfficientContextfreeParsing1970}. The purpose of these
case studies is to showcase possible use cases of resolvable
ambiguity in the context of composing separately defined DSLs (a
subset of Orc, Section~\ref{sec:evaluation-orc}), and of defining
a pre-existing general purpose programming language (OCaml,
Section~\ref{sec:evaluation-ocaml}.

The syntax definition DSL builds on the ``syncons'' introduced by
\citet{palmkvistCreatingDomainSpecificLanguages2019}, wherein each
production is defined separately from each other (one
\syncon{syncon} each).
%
The main difference between the DSL and the formalism in
Section~\ref{sec:parse-time-disambiguation} is that marks are
introduced as separate \syncon{forbid} declarations rather than
inline in productions. This allows supporting high-level
convenience constructs, e.g., specifying precedence between
previously defined operators instead of manually inserting marks,
but is also important for composability as ambiguities can be
statically resolved without changing the original definitions.
Other convenience constructs include defining prefix, infix and
postfix operators with a given associativity.

The running example used in
Section~\ref{sec:parse-time-disambiguation}
(Fig.~\ref{fig:running-example-definition} on
page~\pageref{fig:running-example-definition}) can be defined as
follows:

\begin{tabular}{ll}
\begin{lstlisting}[language=syncon,boxpos=t]
type Exp
grouping "(" Exp ")"
precedence {
  mul; // higher in precedence list
  add; // means higher precedence
}
\end{lstlisting}
&
\begin{lstlisting}[language=syncon,boxpos=t]
token Integer = "[0-9]+"
syncon literal: Exp = n:Integer
syncon list: Exp =
  "[" (head:Exp (";" tail:Exp)*)? "]"
infix add: Exp = "+"
infix mul: Exp = "*"
\end{lstlisting}
\end{tabular}\smallskip

\noindent Here, we define a syntax type \syncon{Exp} for
expressions, and declare that parentheses can be used to group
expressions. We define a lexical token for integers, and a
\syncon{syncon} for integer literals using this token. A list is
defined as a sequence of zero or more expressions separated by
semi-colons wrapped in square braces. Finally, addition and
multiplication is defined as infix syncons with the expected
precedence rules. Note that we could just as well have replaced
the precedence list with explicit \syncon{forbid} declarations,
``\syncon{forbid mul.left = add}'' and ``\syncon{forbid mul.right
  = add}'' (cf. the mark $\{a\}$ in the production of $m$ in
Fig.~\ref{fig:running-example-definition}).

% If marks are needed in non-precedence related
% situations, e.g., adding sequential composition to the running
% example, they can be introduced with ``forbids'':

% \begin{lstlisting}[language=syncon]
% infix seqComp: Exp = ";"
% forbid list.head = seqComp
% forbid list.tail = seqComp
% \end{lstlisting}

\subsection{Composing Language Definitions} \label{sec:evaluation-orc}

In this section, we consider the use case of defining a language
and composing it with another previously defined language, showing
how we can deal with the resulting ambiguities that arise from the
composition.
%
The language being defined is a subset of
Orc~\cite{kitchinOrc2009}, a functional programming language which
includes a number of special-purpose combinators for coordinating
concurrent workflows. On their own, these combinators act as a DSL
for concurrency.

@@

\subsection{Applying Resolvable Ambiguity to a Full Language} \label{sec:evaluation-ocaml}

% \subsection{Cyphym} \label{sec:evaluation-cyphym}
